---
title: "Prediction Assignment Writeup"
author: "Wenzel Vaughn Pestano"
date: "December 2, 2018"
output:
  html_document: default
  
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

The scripts were created and tested on Acer Aspire ES15 and R Studio Version 1.1.383

# Background 

Using devices such as Jawbone Up, Nike FuelBand, and Fitbit it is now possible to collect a large amount of data about personal activity relatively inexpensively. These type of devices are part of the quantified self movement - a group of enthusiasts who take measurements about themselves regularly to improve their health, to find patterns in their behavior, or because they are tech geeks. One thing that people regularly do is quantify how much of a particular activity they do, but they rarely quantify how well they do it. In this project, your goal will be to use data from accelerometers on the belt, forearm, arm, and dumbell of 6 participants. They were asked to perform barbell lifts correctly and incorrectly in 5 different ways. More information is available from the website here: http://web.archive.org/web/20161224072740/http:/groupware.les.inf.puc-rio.br/har (see the section on the Weight Lifting Exercise Dataset).


#Data Sources

The training and test data used for this project are the following:
https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv

https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv

The data for this project come from this source: http://web.archive.org/web/20161224072740/http:/groupware.les.inf.puc-rio.br/har. If you use the document you create for this class for any purpose please cite them as they have been very generous in allowing their data to be used for this kind of assignment.

#Project Objectives
The goal of the project is to predict the manner in which they did the exercise. 
This is the "classe" variable in the training set. You may use any of the other variables to predict with. You should create a report describing how you built your model, how you used cross validation, what you think the expected out of sample error is, and why you made the choices you did. You will also use your prediction model to predict 20 different test cases.


# Preliminary Work
##Reproduceability
Seed was set to 1234 for all code to reproduce the same results.
Different packages were downloaded and installed (e.g. caret, randomForest)

##Building Model

Our outcome variable is classe, a factor variable with 5 levels. For this data set, "participants were asked to perform one set of 10 repetitions of the Unilateral Dumbbell Biceps Curl in 5 different fashions:

exactly according to the specification (Class A)

throwing the elbows to the front (Class B)

lifting the dumbbell only halfway (Class C)

lowering the dumbbell only halfway (Class D)

throwing the hips to the front (Class E)?

Class A corresponds to the specified execution of the exercise, while the other 4 classes correspond to common mistakes." 

Prediction evaluations are based on maximizing the accuracy and minimizing the out-of-sample error. 
All other available variables remained after cleaning will be used for prediction.
Two models will be tested using decision tree and random forest algorithms.
The model with the highest accuracy will be chosen as our final model.

##Packages, Libraries and Seed
Installing packages, loading libraries, and setting the seed for reproduceability:
```{r}
library(caret)
```
```{r}
library(randomForest)
```
```{r}
library(rpart) 
```
```{r}
library(rpart.plot)
```
```{r}
library(RColorBrewer)
```
```{r}
library(rattle)
```
```{r}
set.seed(1234)
```
#Data Getting and Cleaning

```{r}
trainUrl <- "http://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv"
```
```{r}
testUrl <- "http://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv"
```

##Data Loading

```{r}
training <- read.csv(url(trainUrl), na.strings=c("NA","#DIV/0!",""))
testing <- read.csv(url(testUrl), na.strings=c("NA","#DIV/0!",""))
```

##Training Partition
As per rule of thumb, partioning Training data will set into two data sets, 60% for myTraining, 40% for myTesting:
```{r}
TTrain <- createDataPartition(y=training$classe, p=0.6, list=FALSE)
myTraining <- training[TTrain, ]; myTesting <- training[-TTrain, ]
dim(myTraining); dim(myTesting)
```

##Data Cleaning
Transformations are needed to clean the data:

Transform 1: Viewing possible NearZeroVariance (NZV) variables:
```{r}
myDataNZV <- nearZeroVar(myTraining, saveMetrics=TRUE)
```
Subset of NZV variables
```{r}
myNZVvars <- names(myTraining) %in% c("new_window", "kurtosis_roll_belt", "kurtosis_picth_belt",
"kurtosis_yaw_belt", "skewness_roll_belt", "skewness_roll_belt.1", "skewness_yaw_belt",
"max_yaw_belt", "min_yaw_belt", "amplitude_yaw_belt", "avg_roll_arm", "stddev_roll_arm",
"var_roll_arm", "avg_pitch_arm", "stddev_pitch_arm", "var_pitch_arm", "avg_yaw_arm",
"stddev_yaw_arm", "var_yaw_arm", "kurtosis_roll_arm", "kurtosis_picth_arm",
"kurtosis_yaw_arm", "skewness_roll_arm", "skewness_pitch_arm", "skewness_yaw_arm",
"max_roll_arm", "min_roll_arm", "min_pitch_arm", "amplitude_roll_arm", "amplitude_pitch_arm",
"kurtosis_roll_dumbbell", "kurtosis_picth_dumbbell", "kurtosis_yaw_dumbbell", "skewness_roll_dumbbell",
"skewness_pitch_dumbbell", "skewness_yaw_dumbbell", "max_yaw_dumbbell", "min_yaw_dumbbell",
"amplitude_yaw_dumbbell", "kurtosis_roll_forearm", "kurtosis_picth_forearm", "kurtosis_yaw_forearm",
"skewness_roll_forearm", "skewness_pitch_forearm", "skewness_yaw_forearm", "max_roll_forearm",
"max_yaw_forearm", "min_roll_forearm", "min_yaw_forearm", "amplitude_roll_forearm",
"amplitude_yaw_forearm", "avg_roll_forearm", "stddev_roll_forearm", "var_roll_forearm",
"avg_pitch_forearm", "stddev_pitch_forearm", "var_pitch_forearm", "avg_yaw_forearm",
"stddev_yaw_forearm", "var_yaw_forearm")
myTraining <- myTraining[!myNZVvars]

dim(myTraining)
```
Transform 2: Removing ID variable to avoid interference with Machine Learning algorithms
```{r}
myTraining <- myTraining[c(-1)]
```
Transform 3: Removing variables with too many NAs (threshold greater than 60% of NAs)
```{r}
trainingit <- myTraining #subset for looping
for(i in 1:length(myTraining)) { #every column in the training dataset
        if( sum( is.na( myTraining[, i] ) ) /nrow(myTraining) >= .6 ) { #if NAs > 60% of total observations
        for(j in 1:length(trainingit)) {
            if( length( grep(names(myTraining[i]), names(trainingit)[j]) ) ==1)  { #if same columns
                trainingit <- trainingit[ , -j] #Remove that column
            }   
        } 
    }
}
#To check number of observations
dim(trainingit)
```

```{r}
#Reverting to the set:
myTraining <- trainingit
rm(trainingit)
```

Repeating same 3 transformations for myTesting and testing data sets.
```{r}
clean1 <- colnames(myTraining)
clean2 <- colnames(myTraining[, -58]) #already with classe column removed
myTesting <- myTesting[clean1]
testing <- testing[clean2]

#Check the number of observations
dim(myTesting)
```

#To check the new number of observations
```{r}
dim(testing)
```
Coercing Same Data Type for Decision Trees and functioning of Random Forest algorithm with Test dataset

```{r}
for (i in 1:length(testing) ) {
        for(j in 1:length(myTraining)) {
        if( length( grep(names(myTraining[i]), names(testing)[j]) ) ==1)  {
            class(testing[j]) <- class(myTraining[i])
        }      
    }      
}

#TO ensure coercion really worked

testing <- rbind(myTraining[2, -58] , testing)
testing <- testing[-1,]
```

#Using Machine Learning Algorithm with Decision Tree
```{r}
modFitdt <- rpart(classe ~ ., data=myTraining, method="class")
```
Viewing decision tree
```{r}
fancyRpartPlot(modFitdt)
```

#Predicting
```{r}
predictiondt <- predict(modFitdt, myTesting, type = "class")
```

Testing results using confusion matrix 
```{r}
confusionMatrix(predictiondt, myTesting$classe)
```
#Using Machine Learning Algorithm with Random Forests
```{r}
modFitrf <- randomForest(classe ~. , data=myTraining)
```
In-sample error prediction
```{r}
predictionrf <- predict(modFitrf, myTesting, type = "class")
```
Testing results using confusion matrix 
```{r}
confusionMatrix(predictionrf, myTesting$classe)
```

#Conclusion 
Based from the predictions, Random Forests yielded generated better results.

The following statements are the reason why is better predictor than decision trees:

* Random forests are suitable for handling huge number of inputs, especially when the interactions between variables are unknown.
* The built in cross-validation component, giving unbiased estimate of the forest's error rate
* Random forest can also handle unscaled variables and categorical variables which is better for cleaning of data.